Measuring contrast processing in the visual system using SSVEP

### 2024-11-28 review.qmd
Tidying

Similar changes to the contrast response function might also be obtained using adaptation paradigms, in which the visual system is exposed to high contrast stimuli for long durations. Psychophysically, adaptation increases detection thresholds, but has little effect on contrast discrimination performance [@Ross1993], much like pattern masks [@Foley1994]. Although SSVEP adaptation effects show strong tuning for orientation [@Vergeer2018; @Campbell1970] and spatial frequency [@Mecacci1976], there appear to be very measurements of the full SSVEP contrast response function before and after an extended period of adaptation. One exception is a study by Bach et al [@BachGreenleeBuhler1988] that examined the effects of adaptation on SSVEP responses who found a paradoxical increase in the 1F SSVEP amplitude following prolonged exposure to high-contrast stimuli across much of the probe contrast range. They explain this as being due to the presence of a nonlinear transducer function that saturates at high probe contrast driving the response to higher harmonic frequencies. Adaptation reduces the apparent contrast of the input, leading to less distortion, a more sinusoidal output and a concentration of the frequency-tagged response at lower harmonics. This explanation is conceptually related to the 'supersaturation' that we observe in Figure xxx.

@@

A robust estimate of the threshold therefore requires the measurement of the SSVEP amplitude at many different super-threshold contrast levels. This was made faster by the development of the ‘sweep VEP’ paradigm in which the stimulus changed its contrast, spatial frequency or some other property, throughout a trial [@Tyler1979]. To avoid hysteresis effects, the sweep is sometimes conducted both up and down in the same experiment [@Tyler1979; @Norcia1985a; @Norcia1990]. The sweep VEP (really, a sweep 'SSVEP') technique is now commonly-used to obtain a rapid and objective measurement of visual acuity. In particular, because of its relative speed and simplicity, this technique has now become a standard for conducting tests of visual acuity in very young subjects or where behavioural tests are not appropriate [@Ridder2004; @Hoffmann2017; @Bach2008; @Bach2020].

@@

The SSVEP technique has also been used to study the development of the contrast gain control mechanisms described in the previous section [@Candy2001; @Pei2017]. Although contrast gain control is measurable in infants as young as six weeks old [@Candy2001; @Morrone1986; @Skoczenski1998], its development appears to be slower, with adult levels being reached at approximately 11 years [@Pei2017].

>>>>

Similar changes to the contrast response function might also be obtained using adaptation paradigms, in which the visual system is exposed to high contrast stimuli for long durations. Psychophysically, adaptation increases detection thresholds, but has little effect on contrast discrimination performance [@Ross1993], much like pattern masks [@Foley1994]. Although SSVEP adaptation effects show strong tuning for orientation [@Vergeer2018; @Campbell1970] and spatial frequency [@Mecacci1976], there appear to be very measurements of the full SSVEP contrast response function before and after an extended period of adaptation. One exception is a study by Bach et al [@BachGreenleeBuhler1988] that examined the effects of adaptation on SSVEP responses who found a paradoxical increase in the 1F SSVEP amplitude following prolonged exposure to high-contrast stimuli across much of the probe contrast range. They explain this as being due to the presence of a nonlinear transducer function that saturates at high probe contrast driving the response to higher harmonic frequencies. Adaptation reduces the apparent contrast of the input, leading to less distortion, a more sinusoidal output and a concentration of the frequency-tagged response at lower harmonics. This explanation is conceptually related to the 'supersaturation' effect that we observe in Figure 4.

@@

A robust estimate of the threshold therefore requires the measurement of the SSVEP amplitude at many different super-threshold contrast levels. This was made faster by the development of the ‘sweep VEP’ paradigm in which the stimulus changed its contrast, spatial frequency or some other property, throughout a trial [@Tyler1979]. To avoid hysteresis or short-term adaptation effects, the sweep is sometimes conducted both up and down in the same experiment [@Tyler1979; @Norcia1985a; @Norcia1990]. The sweep VEP (really, a sweep 'SSVEP') technique is now commonly-used to obtain a rapid and objective measurement of visual acuity. In particular, because of its relative speed and simplicity, this technique has now become a standard for conducting tests of visual acuity in very young subjects or where behavioural tests are not appropriate [@Ridder2004; @Hoffmann2017; @Bach2008; @Bach2020].

@@

The SSVEP technique has also been used to study the development of the contrast gain control mechanisms described in the previous section [@Candy2001; @Pei2017]. Although response- [@Morrone1986] or contrast gain control [@Candy2001; ; @Skoczenski1998] is measurable in infants as young as six weeks old, its development appears to be slower, with adult levels being reached at approximately 11 years [@Pei2017].
